---
title: "Permafrost mixing study"
subtitle: "Amplicon data: microbial community analyses"
author: "Stacey Doherty - adapted from Baker et al. 2023 doi: 10.1088/1748-9326/acc542."
date: "07/02/2024"
output:
    html_document:
        toc: true
editor_options: 
  chunk_output_type: inline
---

<!-- Render with rmarkdown::render('code/pf_mixing_amplicon_analysis.Rmd', output_file = '../analysis/pf_mixing_amplicon_analysis.html') -->

<!-- More analysis resources on phyloseq website: https://joey711.github.io/phyloseq -->

<!-- Accessor functions: https://joey711.github.io/phyloseq/preprocess.html -->
<!-- https://bioconductor.riken.jp/packages/3.0/bioc/vignettes/phyloseq/inst/doc/phyloseq-basics.html#table:access -->

```{r general settings, include=FALSE}
knitr::opts_chunk$set(
    message = FALSE,
    warning = FALSE,
    echo = FALSE,
    fig.path = "../figures/amplicon/",
    # fig.align='center',
    fig.width = 11,
    out.width = "100%",
    dev = c("png", "pdf")
)

# include = FALSE --> execute but do not print code or results
# echo = FALSE --> execute but do not print code
# results = "hide" --> do not print results
# eval = FALSE --> do not evaluate
# warning = FALSE and message = FALSE --> hide warnings or messages produced
# fig.height and fig.width --> control figure size in inches
# out.width = "xx%" --> control figure size as fraction of HTML display width

# theme_set(theme_bw())
```

```{r load libraries, include=FALSE}

library("here")
library("tidyverse")
library("scales")
# library("ggbeeswarm") # for geom_quasirandom

library("cowplot") # to align multiple plots

library("patchwork") # to align multiple plots - use for the ordinations to retain fixed aspect ratio

# note: phyloseq install instructions at https://joey711.github.io/phyloseq/install.html
#   use BiocLite rather than BiocManager as used here; the latter is preferred with R > 3.5
# install.packages("BiocManager")
# library("BiocManager")
# BiocManager::install("phyloseq")
library("phyloseq")

#library("Biostrings")

library("vegan")
```

```{r get data}

# loads list of phyloseq objects: one for 16S and one for ITS

    ps.list  <- readRDS(here("/path_to_rds_file", "amplicon.rds"))

# parse metadata

    process_sample_data <- function (sample.data) {
        
        processed.sample.data = sample.data %>% as("data.frame") %>%
            # filter(Project == "permafrost_mix") %>% 
            separate(Description, into = c("site", "mix_ratio", "replicate", "timepoint"), sep = "-", remove = FALSE) %>%
            mutate(mix_time = paste(mix_ratio, timepoint, sep = "-")) %>%
            mutate(site = factor(site, levels = c("APT", "TK1"))) %>%
            mutate(mix_ratio = factor(mix_ratio, levels = c("100AL", "10PF_90AL" ,"50PF_50AL", "90PF_10AL", "100PF", "100TZ"))) %>%
            mutate(replicate = factor(gsub("^R","", x = replicate), levels = c("1","2","3","4"))) %>%
            mutate(timepoint = factor(timepoint, levels = c("t0","tf"))) %>%
            mutate(mix_time = factor(mix_time, levels = c("100AL-t0", "100AL-tf", "10PF_90AL-tf" ,"50PF_50AL-tf", "90PF_10AL-tf", "100PF-t0", "100PF-tf", "100TZ-t0", "100TZ-tf"))) %>%
            select(-Project, -BarcodePlate, -Barcode.Well)
    
        return(processed.sample.data) # yields a data frame
        
    }

    for (dataset in c("16S","ITS")) {
    
        sample_data(ps.list[[dataset]]) <- process_sample_data(sample_data(ps.list[[dataset]]))
      
        # add read counts to metadata
        sample_data(ps.list[[dataset]])$Reads <- ps.list[[dataset]] %>% sample_sums()
    
    }

```

```{r extract metadata to tibble}

sample.data <- bind_rows(
    ps.list$`16S` %>% sample_data() %>% as("data.frame"),
    ps.list$`ITS` %>% sample_data() %>% as("data.frame")
)

rownames(sample.data) <- NULL
  
```

**************************

## Choosing rarefaction depth

### Samples ranked by read count

```{r view read counts and choose rarefaction depths}
  
# order samples by read counts

    sample.data.ranked.by.reads <- sample.data %>%
        group_by(Dataset) %>%
        mutate(Rank = as.integer(rank(`Reads`))) %>% 
        arrange(`Reads`, .by_group = TRUE) %>%
        ungroup()

# empty list for rarefaction depths

    rarefaction.depth <- list("16S" = NA, "ITS" = NA)

```

Here are plots of read counts in all samples, ranked in order of read count, with samples colored by different variables. 

```{r read_counts_ranked_samples, fig.width=8, out.width="73%", results='asis'}
    
# samples in order of read count

# pdf(here("figures","read count ranked samples.pdf"), width = 8, height = 4, useDingbats = FALSE)

#set colors for plots
hex_colors <- c("#d73027", "#fc8d59", "#fee090", "#91bfdb", "#4575b4", "#91cf60") 

    # read counts colored by different variables
  
    read.count.plot <- sample.data.ranked.by.reads %>%
        ggplot() + geom_point(aes(x = Rank, y=Reads)) +
            facet_wrap(~ Dataset, scales = "free", nrow = 1) + 
            scale_y_continuous(labels = label_number(suffix = " K", scale = 1e-3)) +
            expand_limits(y = 0) + scale_color_manual(values = hex_colors)
    
    read.count.plot + aes(col = site) + theme(legend.title = element_blank())
    cat("</br>")

    read.count.plot + aes(col = mix_ratio) + theme(legend.title = element_blank())
    cat("</br>")
  
    read.count.plot + aes(col = timepoint)
    cat("</br>")
    
    rm(read.count.plot)
  
#graphics.off()
 
```

### Read count boxplots

Here are boxplots of read counts for all samples broken down by different variables. 

```{r read_count_boxplots, fig.width = 10, fig.height = 4, out.width="36%", results="asis"}

#pdf(here("figures","read count boxplots.pdf"), width = 8, height = 4, useDingbats = FALSE)

    # read counts colored by different variables
  
    read.count.boxplot <- sample.data %>%
        ggplot() + geom_boxplot(aes(y=Reads)) +
            facet_wrap(~ Dataset, scales = "free", nrow = 1) + 
            scale_y_continuous(labels = label_number(suffix = " K", scale = 1e-3)) +
            expand_limits(y = 0) + scale_color_manual(values = hex_colors)+
            theme(legend.position = "bottom")
    
    read.count.boxplot + aes(x = site) + theme(axis.title.x = element_blank())
    cat("</br>")
  
    read.count.boxplot + aes(x = mix_ratio) + theme(axis.title.x = element_blank())
    cat("</br>")
  
    read.count.boxplot + aes(x = timepoint)
    cat("</br>")

    rm(read.count.boxplot)
  
#graphics.off()


```

### Choice of rarefaction depth

The plots above suggest a rarefaction depth of 10000 reads might work? Here is where that depth sits relative to the samples in each dataset.

```{r read_counts_and_rarefaction_depths, fig.width=8, out.width="73%"}

rarefaction.depth <- list(
    "16S" = 10000,
    "ITS" = 10000
)

# pdf(here("figures","read counts with rarefaction depths.pdf"), width = 6, height = 4, useDingbats = FALSE)

    rarefaction.depth.df <- data.frame(Reads = do.call("c", rarefaction.depth)) %>% rownames_to_column(var = "Dataset")

    sample.data.ranked.by.reads %>%
        ggplot() + geom_point(aes(x = Rank, y = Reads)) +
            facet_wrap(~ Dataset, scales = "free") + 
            scale_y_continuous(labels = label_number(suffix = " K", scale = 1e-3)) +
            expand_limits(y = 0) + scale_color_manual(values = hex_colors) +
            geom_hline(data = rarefaction.depth.df, aes(yintercept = Reads), lty = "dashed", col = "red") +
            geom_text(data = rarefaction.depth.df, aes(y = Reads, label = paste("rarefaction depth =", Reads)), x = 50, col = "red", vjust = -0.5)

# graphics.off()        
```

## Alpha diversity analysis

```{r prepare alpha diversity data}

# remove extraction blanks from downstream analysis

ps.list_filtered <- list(
        "16S" = subset_samples(ps.list[["16S"]], !is.na(site)),
        "ITS" = subset_samples(ps.list[["ITS"]], !is.na(site))
    )

# a single rarefaction for each dataset
# - probably need to repeat this if we are going to publish

    ps.list.rare.1 <- list(
        "16S" = rarefy_even_depth(ps.list_filtered[["16S"]],
            sample.size = rarefaction.depth[["16S"]],
            rngseed = 123, replace = TRUE, trimOTUs = TRUE, verbose = TRUE),
        "ITS" = rarefy_even_depth(ps.list_filtered[["ITS"]],
            sample.size = rarefaction.depth[["ITS"]],
            rngseed = 123, replace = TRUE, trimOTUs = TRUE, verbose = TRUE)
    )

# add diversity measures

    for (dataset in c("16S","ITS")) {
        sample_data(ps.list.rare.1[[dataset]])$richness <- otu_table(ps.list.rare.1[[dataset]]) %>%
            apply(MARGIN = 1, FUN = function (X) sum(X > 0))
        sample_data(ps.list.rare.1[[dataset]])$shannon <- vegan::diversity(otu_table(ps.list.rare.1[[dataset]]),
            index = "shannon", MARGIN = 1)
        sample_data(ps.list.rare.1[[dataset]])$invsimp <- vegan::diversity(otu_table(ps.list.rare.1[[dataset]]),
            index = "invsimpson", MARGIN = 1)
    }

# extract diversity results data for plotting

    alpha.div.data <- bind_rows(
        "16S" = sample_data(ps.list.rare.1[["16S"]]) %>% as("data.frame"),
        "ITS" = sample_data(ps.list.rare.1[["ITS"]]) %>% as("data.frame")
    ) %>%
    rename(`Observed_ASV_richness` = richness) %>%
    rename(`Shannon_diversity` = shannon) %>%
    rename(`Inverse_Simpson_diversity` = invsimp)


```

### Species richness plots

```{r alpha_diversity_richness, fig.width=8, out.width="73%"}
# pdf(here("figures","alpha diversity - observed richness.pdf"), width = 8, height = 4, useDingbats = FALSE)

    for (dataset in c("16S","ITS")) {

            alpha.plot <- alpha.div.data %>% filter(Dataset == dataset) %>%
                ggplot(aes(x = `mix_ratio`, y = `Observed_ASV_richness`, col = factor(mix_ratio))) +
                    geom_point() + facet_grid(. ~ site + timepoint, scales = "free_x", space = "free") +
                    ggtitle(dataset) +
                    expand_limits(y=0) + theme(legend.title = element_blank()) +
                    theme_bw() +
                    scale_color_manual(values = hex_colors, na.value = "grey")
                    
            print(alpha.plot)
            
            rm(alpha.plot)
    
    }  

# graphics.off()

```

```{r alpha_diversity_richness, fig.width=8, out.width="73%"}
# pdf(here("figures","alpha diversity - observed richness.pdf"), width = 8, height = 4, useDingbats = FALSE)

    for (dataset in c("16S","ITS")) {

            alpha.plot <- alpha.div.data %>% filter(Dataset == dataset) %>%
                ggplot(aes(x = `mix_time`, y = `Observed_ASV_richness`, col = factor(mix_ratio))) +
                    geom_boxplot() + facet_grid(. ~ site + timepoint, scales = "free_x", space = "free") +
                    ggtitle(dataset) +
                    expand_limits(y=0) + theme(legend.title = element_blank()) +
                    theme_bw() +
                    scale_color_manual(values = hex_colors, na.value = "grey")
                    
            print(alpha.plot)
            
            rm(alpha.plot)
    
    }  

# graphics.off()

```



### Shannon diversity plots

```{r alpha_diversity_shannon, fig.width=8, out.width="73%"}

# pdf(here("figures","alpha diversity - shannon diversity.pdf"), width = 8, height = 4, useDingbats = FALSE)

    for (dataset in c("16S","ITS")) {

            alpha.plot <- alpha.div.data %>% filter(Dataset == dataset) %>%
                ggplot(aes(x = `mix_time`, y = `Shannon_diversity`, col = factor(mix_ratio))) +
                    geom_point() + facet_grid(. ~ site + timepoint, scales = "free_x", space = "free")+
                    ggtitle(dataset) +
                    expand_limits(y=0) + theme(legend.title = element_blank()) + theme_bw() +
                    scale_color_manual(values = hex_colors, na.value = "grey")
                    
            print(alpha.plot)
            
            rm(alpha.plot)
    
    }  

# graphics.off()
```

```{r alpha_diversity_shannon, fig.width=8, out.width="73%"}

# pdf(here("figures","alpha diversity - shannon diversity.pdf"), width = 8, height = 4, useDingbats = FALSE)

    for (dataset in c("16S","ITS")) {

            alpha.plot <- alpha.div.data %>% filter(Dataset == dataset) %>%
                ggplot(aes(x = `mix_time`, y = `Shannon_diversity`, col = factor(mix_ratio))) +
                    geom_boxplot() + facet_grid(. ~ site + timepoint, scales = "free_x", space = "free") +
                    ggtitle(dataset) +
                    expand_limits(y=0) + theme_bw() + theme(
                      text = element_text(size = 18),
                      axis.title = element_text(size = 14),
                      axis.text = element_text(size = 12),
                      legend.title = element_text(size = 14),
                      legend.text = element_text(size = 12)) +
                    scale_color_manual(values = hex_colors, na.value = "grey")
                    
            print(alpha.plot)
            
            rm(alpha.plot)
    
    }  

# graphics.off()
```

### Alpha diversity stats

```{r prepare data by subsetting sites and timepoints}

#remove ITS data
alpha.div.data.16S <- subset(alpha.div.data, Dataset == "16S")

#subset data into sites for anova
div_tf_apt <- subset(alpha.div.data.16S, site == "APT" & timepoint == "tf")
div_tf_tk1 <- subset(alpha.div.data.16S, site == "TK1" & timepoint == "tf")
div_t0_apt <- subset(alpha.div.data.16S, site == "APT" & timepoint == "t0")
div_t0_tk1 <- subset(alpha.div.data.16S, site == "TK1" & timepoint == "t0")

################  ANOVA Shannon diversity
######## Compute the analysis of variance for all dataframes above. Note only one example is shown here 
div_apt.aov <- aov(Shannon_diversity ~ mix_ratio, data = div_t0_apt)

## Summary of the analysis
summary(div_apt.aov)

######### Check Assumptions of the test

## Assumption 1: Normality of the residual
plot(div_apt.aov,2) #Q-Q plot

shapiro.test(x = residuals(object = div_apt.aov)) # Shapiro-Wilk test
## data are  NORMAL

## Assumption 2: Homogeineity of Variance
library(car) # Use Levene test from the car library

leveneTest(Shannon_diversity ~ mix_ratio, data = div_t0_apt)
## data do  have common variance

#tukey multiple comparison

library(multcomp)
library(multcompView)

my.mod = lm(Shannon_diversity ~ mix_ratio, data = div_t0_apt)
my.tukey <- TukeyHSD(aov(my.mod))

my.letters <- multcompLetters4(my.mod, my.tukey, reversed = TRUE)
as.data.frame.list(my.letters$mix_ratio)[1]

```



## Beta diversity analysis


```{r prepare data by subsetting sites}

# remove extraction blanks from downstream analysis

ps.list_filtered_apt <- list(
        "16S" = subset_samples(ps.list_filtered[["16S"]], site != "TK1"),
        "ITS" = subset_samples(ps.list_filtered[["ITS"]], site != "TK1")
    )

ps.list_filtered_tk1 <- list(
        "16S" = subset_samples(ps.list_filtered[["16S"]], site != "APT"),
        "ITS" = subset_samples(ps.list_filtered[["ITS"]], site != "APT")
    )

```

# beta diversity - all samples
```{r calculate average bray-curtis distances}

# avgdist rarefaction for beta diversity analysis

# avgdist(x, sample, distfun = vegdist, meanfun = mean, transf = NULL, iterations = 100, dmethod = "bray", ...)

## N.B. currently only running 10 iterations ##
## increase this value for final analysis ##

num.iter <- 1000 #change to 10 for prelim analysis

bray.dists <- list(
    "16S" = ps.list_filtered[["16S"]] %>% otu_table() %>% as("matrix") %>% 
        avgdist(sample = rarefaction.depth[["16S"]], iterations = num.iter, dmethod = "bray"),
    "ITS" = ps.list_filtered[["ITS"]] %>% otu_table() %>% as("matrix") %>% 
        avgdist(sample = rarefaction.depth[["ITS"]], iterations = num.iter, dmethod = "bray")
)

saveRDS(bray.dists, file = here("bray_dists_1000.rds"))

# or load up previous data:

#bray.dists <- readRDS(file = here("bray_dists_1000.rds"))
    
```

```{r calculate ordinations, include = FALSE}
# ordinations for rarefied OTU tables

    bray_curtis_NMDS <- lapply(bray.dists, metaMDS)

```      

```{r NMDS_plots, fig.width = 8, out.width="73%"}

# pdf(here("figures","beta diversity - bray curtis NMDS.pdf"), width = 6, height = 5, useDingbats = FALSE)

    nmds.plot <- list()

    for (dataset in c("16S","ITS")) {
        
        my.title.text <- paste(dataset, "Bray-Curtis")
        
        nmds.plot[[dataset]] <- plot_ordination(ps.list_filtered[[dataset]], bray_curtis_NMDS[[dataset]], type = "samples") +
            coord_equal() + theme_bw() + theme(legend.position = "bottom") + ggtitle(my.title.text) + geom_point(size = 2.5) +
            annotate("text", x = Inf, y = Inf, hjust = 1.25, vjust = 2.5,
                label = paste0("stress = ", sprintf("%.2f", bray_curtis_NMDS[[dataset]]$stress)))

        rm(my.title.text)

    }

    # plot ordinations using patchwork package to retain fixed aspect ratio
    #  - the "+" between plots indicates to align them 
  
    # color by site treatment 
    (nmds.plot[["16S"]] + aes(color = site) + theme(legend.title = element_blank())) +
        (nmds.plot[["ITS"]] + aes(color = site) + theme(legend.title = element_blank()))

    # color by mix_ratio treatment 
    (nmds.plot[["16S"]] + aes(color = mix_ratio, shape = timepoint) + theme(legend.title = element_blank()) + scale_color_manual(values = hex_colors, na.value = "grey")) +
        (nmds.plot[["ITS"]] + aes(color = mix_ratio, shape = timepoint) + theme(legend.title = element_blank()) + scale_color_manual(values = hex_colors, na.value = "grey"))

    # color by timepoint treatment 
    (nmds.plot[["16S"]] + aes(color = timepoint) + theme(legend.title = element_blank())) +
        (nmds.plot[["ITS"]] + aes(color = timepoint) + theme(legend.title = element_blank()))

    # add one where mix_ratio is color and shape is timepoint
      
# graphics.off()

```



# beta diversity - apt samples
```{r calculate average bray-curtis distances}

# avgdist rarefaction for beta diversity analysis

# avgdist(x, sample, distfun = vegdist, meanfun = mean, transf = NULL, iterations = 100, dmethod = "bray", ...)

## N.B. currently only running 10 iterations ##
## increase this value for final analysis ##

num.iter <- 1000

bray.dists <- list(
    "16S" = ps.list_filtered_apt[["16S"]] %>% otu_table() %>% as("matrix") %>% 
        avgdist(sample = rarefaction.depth[["16S"]], iterations = num.iter, dmethod = "bray"),
    "ITS" = ps.list_filtered_apt[["ITS"]] %>% otu_table() %>% as("matrix") %>% 
        avgdist(sample = rarefaction.depth[["ITS"]], iterations = num.iter, dmethod = "bray")
)

saveRDS(bray.dists, file = here("bray_dists_1000_apt.rds"))

# or load up previous data:

#bray.dists <- readRDS(file = here("bray_dists_1000_apt.rds"))

#Permanova 
sample_data_df_16 <- ps.list_filtered_apt[["16S"]] %>% sample_data() %>% as("data.frame")
sample_data_df_16 <- sample_data_df_16[sample_data_df_16$SampleID != "16S_APT.10PF_90AL.R4.tf", ] #remove samples that were dropped after rarefying
adonis_16S <- adonis2(bray.dists[["16S"]]~ mix_ratio, data=sample_data_df_16, by=NULL) 

    
```

```{r calculate ordinations, include = FALSE}
# ordinations for rarefied OTU tables

    bray_curtis_NMDS <- lapply(bray.dists, metaMDS)

```      

```{r NMDS_plots, fig.width = 8, out.width="73%"}

# pdf(here("figures","beta diversity - bray curtis NMDS.pdf"), width = 6, height = 5, useDingbats = FALSE)

    nmds.plot <- list()

    for (dataset in c("16S","ITS")) {
        
        my.title.text <- paste(dataset, "APT Bray-Curtis")
        
        nmds.plot[[dataset]] <- plot_ordination(ps.list_filtered_apt[[dataset]], bray_curtis_NMDS[[dataset]], type = "samples") +
            coord_equal() + theme_bw() + theme(legend.position = "right") + ggtitle(my.title.text) + geom_point(size = 2.5) +
            annotate("text", x = Inf, y = Inf, hjust = 1.25, vjust = 2.5,
                label = paste0("stress = ", sprintf("%.2f", bray_curtis_NMDS[[dataset]]$stress)))

        rm(my.title.text)

    }

    # plot ordinations using patchwork package to retain fixed aspect ratio
    #  - the "+" between plots indicates to align them 

    # color by mix_ratio treatment 
    (nmds.plot[["16S"]] + aes(color = mix_ratio, shape = timepoint) + theme(legend.title = element_blank()) + scale_color_manual(values = hex_colors, na.value = "grey")) +
        (nmds.plot[["ITS"]] + aes(color = mix_ratio, shape = timepoint) + theme(legend.title = element_blank()) + scale_color_manual(values = hex_colors, na.value = "grey"))

      
# graphics.off()

```



# beta diversity - tk1 samples
```{r calculate average bray-curtis distances}

# avgdist rarefaction for beta diversity analysis

# avgdist(x, sample, distfun = vegdist, meanfun = mean, transf = NULL, iterations = 100, dmethod = "bray", ...)

## N.B. currently only running 10 iterations ##
## increase this value for final analysis ##

num.iter <- 1000

bray.dists <- list(
    "16S" = ps.list_filtered_tk1[["16S"]] %>% otu_table() %>% as("matrix") %>% 
        avgdist(sample = rarefaction.depth[["16S"]], iterations = num.iter, dmethod = "bray"),
    "ITS" = ps.list_filtered_tk1[["ITS"]] %>% otu_table() %>% as("matrix") %>% 
        avgdist(sample = rarefaction.depth[["ITS"]], iterations = num.iter, dmethod = "bray")
)

saveRDS(bray.dists, file = here("bray_dists_1000_tk1.rds"))

# or load up previous data:

#bray.dists <- readRDS(file = here("bray_dists_1000_tk1.rds"))


#Permanova 
sample_data_df_16 <- ps.list_filtered_tk1[["16S"]] %>% sample_data() %>% as("data.frame")
sample_data_df_16 <- sample_data_df_16[sample_data_df_16$SampleID != "16S_TK1.100PF.R4.t0", ] #remove samples that were dropped after rarefying
sample_data_df_16 <- sample_data_df_16[sample_data_df_16$SampleID != "16S_TK1.90PF_10AL.R4.tf", ] #remove samples that were dropped after rarefying
adonis_16S <- adonis2(bray.dists[["16S"]]~ mix_ratio, data=sample_data_df_16, by=NULL) 

    
```

```{r calculate ordinations, include = FALSE}
# ordinations for rarefied OTU tables

    bray_curtis_NMDS <- lapply(bray.dists, metaMDS)

```      

```{r NMDS_plots, fig.width = 8, out.width="73%"}

# pdf(here("figures","beta diversity - bray curtis NMDS.pdf"), width = 6, height = 5, useDingbats = FALSE)

    nmds.plot <- list()

    for (dataset in c("16S","ITS")) {
        
        my.title.text <- paste(dataset, "TK1 Bray-Curtis")
        
        nmds.plot[[dataset]] <- plot_ordination(ps.list_filtered_tk1[[dataset]], bray_curtis_NMDS[[dataset]], type = "samples") +
            coord_equal() + theme_bw() + theme(legend.position = "right") + ggtitle(my.title.text) + geom_point(size = 2.5) + 
            annotate("text", x = Inf, y = Inf, hjust = 3.10, vjust = 2.5,
                label = paste0("stress = ", sprintf("%.2f", bray_curtis_NMDS[[dataset]]$stress)))

        rm(my.title.text)

    }

    # plot ordinations using patchwork package to retain fixed aspect ratio
    #  - the "+" between plots indicates to align them 

    # color by mix_ratio treatment 
    (nmds.plot[["16S"]] + aes(color = mix_ratio, shape = timepoint) + theme(legend.title = element_blank()) + scale_color_manual(values = hex_colors, na.value = "grey")) +
        (nmds.plot[["ITS"]] + aes(color = mix_ratio, shape = timepoint) + theme(legend.title = element_blank()) + scale_color_manual(values = hex_colors, na.value = "grey"))

 
# graphics.off()

```


## Taxon barplots

```{r prepare taxon barplot data}
    
# aggregate at phylum (16S) or class (ITS)

    ps.list.rare.1.glom <- list(
            "16S" = ps.list.rare.1[["16S"]] %>% tax_glom("silva_phylum"),
            "ITS" = ps.list.rare.1[["ITS"]] %>% tax_glom("unite_class")
        )

# prune phyloseq objects to top 10 taxa

    top10.names <- lapply(ps.list.rare.1.glom, FUN = function (X) {
        names(sort(taxa_sums(X), decreasing=TRUE))[1:10]
    })

    top10.phyloseq <- sapply(names(top10.names), FUN = function (X) {
        prune_taxa(top10.names[[X]], ps.list.rare.1.glom[[X]])
    })

# get (aggregated) ASV table, taxon metadata, sample metadata for top 10 taxa
    
    top10.phyloseq.otu <- lapply(top10.phyloseq, FUN = function (X) {
        otu_table(X) %>% as.data.frame() %>% rownames_to_column("SampleID")
    })
      
    top10.phyloseq.tax <- list(
      "16S" = tax_table(top10.phyloseq[["16S"]]) %>% as.data.frame() %>% rownames_to_column("ASV") %>% select(ASV, silva_phylum),
      "ITS" = tax_table(top10.phyloseq[["ITS"]]) %>% as.data.frame() %>% rownames_to_column("ASV") %>% select(ASV, unite_class)
    )

    top10.phyloseq.sample <- lapply(top10.phyloseq, FUN = function (X) {
        sample_data(X) %>% as("data.frame") %>% select(SampleID, site, mix_ratio, timepoint, mix_time, replicate)
            # %>% mutate(RDX = factor(ifelse(RDX == "RDX", "+RDX", ifelse(RDX == "No RDX", "-RDX", NA)), levels = c("-RDX", "+RDX"))) %>%
            # mutate(strain.11Y = factor(ifelse(strain.11Y == "11Y", "+11Y", ifelse(strain.11Y == "No 11Y", "-11Y", NA)), levels = c("-11Y", "+11Y"))) %>%
            # mutate(Compost = factor(ifelse(Compost == "Compost", "+Comp", ifelse(Compost == "No compost", "-Comp", NA)), levels = c("-Comp", "+Comp")))
    })

# create dataframe for plotting and clean up names
    
    top10.df <- sapply(names(top10.names), FUN = function (X) {
        full_join(top10.phyloseq.sample[[X]], top10.phyloseq.otu[[X]], by = "SampleID") %>%
            pivot_longer(-c(SampleID, site, mix_ratio, timepoint, mix_time, replicate), names_to = "ASV", values_to = "Rarefied reads") %>%
            left_join(top10.phyloseq.tax[[X]], by = "ASV") %>% select(-ASV)
    }, simplify = FALSE)

    top10.df$`16S` <- top10.df$`16S` %>%
        rename(`Silva phylum` = silva_phylum)

    top10.df$`ITS` <- top10.df$`ITS` %>%
        rename(`Unite class` = unite_class) %>%
        mutate(`Unite class` = gsub("c__", "", `Unite class`))

```


```{r 16S barplot}

# 16S barplots

barplot_silva_phylum <- top10.df[["16S"]] %>%
    ggplot(aes(x = SampleID, y= `Rarefied reads`, fill = `Silva phylum`)) +
        geom_col(width=0.75, position = "stack", col = "black", linewidth = 0) + 
        scale_fill_brewer(palette = "Paired") + theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
        ggtitle ("16S") + labs(x = "Mix Ratio") +
        facet_grid(. ~ site + timepoint, scales = "free_x", space = "free") +
        scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

barplot_silva_phylum

# ggsave(filename = here("figures","barplot_16S_silva_phylum.pdf"), width = 12, height = 7, useDingbats = FALSE)

```

```{r ITS barplot}

# ITS barplots

barplot_unite_class <- top10.df[["ITS"]] %>%
    ggplot(aes(x = SampleID, y = `Rarefied reads`, fill = `Unite class`)) +
        geom_col(width=0.75, position = "stack", col = "black", linewidth = 0) + 
        scale_fill_brewer(palette = "Paired") + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
        ggtitle ("ITS") + labs(x = "Mix Ratio") +
        facet_grid(. ~ site + timepoint, scales = "free_x", space = "free") +
        scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))

barplot_unite_class

# ggsave(filename = here("figures","barplot_ITS_unite_class.pdf"), width = 12, height = 7, useDingbats = FALSE)
```

```{r